
# Data visualization methods in Seurat

https://satijalab.org/seurat/articles/visualization_vignette.html

```{r Load dataset}
SeuratData::InstallData("pbmc3k")

library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
data("pbmc3k.final")
pbmc3k.final$groups <- sample(c("group1", "group2"), size = ncol(pbmc3k.final), replace = TRUE)
features <- c("LYZ", "CCL5", "IL32", "PTPRCAP", "FCGR3A", "PF4")
pbmc3k.final
```

## Five visualizations of marker feature expression

```{r Ridge plot, fig.width=6, fig.height=6}
# Ridge plots - from ggridges. Visualize single cell expression distributions in each cluster
RidgePlot(pbmc3k.final, features = features, ncol = 2)
```

```{r Violin plot, fig.width=6, fig.height=6}
# Violin plot - Visualize single cell expression distributions in each cluster
VlnPlot(pbmc3k.final, features = features)
```

```{r Feature plot, fig.width=6, fig.height=6}
# Feature plot - visualize feature expression in low-dimensional space
FeaturePlot(pbmc3k.final, features = features)
```

```{r Dot plots, fig.width=3, fig.height=2}
# Dot plots - the size of the dot corresponds to the percentage of cells expressing the
# feature in each cluster. The color represents the average expression level
DotPlot(pbmc3k.final, features = features) + RotatedAxis()
```

```{r Heatmap}
# Single cell heatmap of feature expression
DoHeatmap(subset(pbmc3k.final, downsample = 100), features = features, size = 3)
```

## New additions to `FeaturePlot`

```{r Customizing legends, fig.width=3, fig.height=2}
# Plot a legend to map colors to expression levels
FeaturePlot(pbmc3k.final, features = "MS4A1")

# Adjust the contrast in the plot
FeaturePlot(pbmc3k.final, features = "MS4A1", min.cutoff = 1, max.cutoff = 3)
```

```{r Quantile cutoffs, fig.width=4, fig.height=2}
# Calculate feature-specific contrast levels based on quantiles of non-zero expression.
# Particularly useful when plotting multiple markers
FeaturePlot(pbmc3k.final, features = c("MS4A1", "PTPRCAP"), min.cutoff = "q10", max.cutoff = "q90")
```


```{r Coexpression, fig.width=6, fig.height=2}
# Visualize co-expression of two features simultaneously
FeaturePlot(pbmc3k.final, features = c("MS4A1", "CD79A"), blend = TRUE)
```

```{r Expression by group, fig.width=3, fig.height=3}
# Split visualization to view expression by groups (replaces FeatureHeatmap)
FeaturePlot(pbmc3k.final, features = c("MS4A1", "CD79A"), split.by = "groups")
```

## Updated and expanded visualization functions
```{r Split / grouped violin plots, fig.width=4, fig.height=3}
# Violin plots can also be split on some variable. Simply add the splitting variable to object
# metadata and pass it to the split.by argument
VlnPlot(pbmc3k.final, features = "percent.mt", split.by = "groups")
```

```{r Split dot plot, fig.width=3, fig.height=3}
# SplitDotPlotGG has been replaced with the `split.by` parameter for DotPlot
DotPlot(pbmc3k.final, features = features, split.by = "groups") + RotatedAxis()
```

```{r DimPlot - umap, tsne, pca}
# DimPlot replaces TSNEPlot, PCAPlot, etc. In addition, it will plot either 'umap', 'tsne', or
# 'pca' by default, in that order
DimPlot(pbmc3k.final)

pbmc3k.final.no.umap <- pbmc3k.final
pbmc3k.final.no.umap[["umap"]] <- NULL
DimPlot(pbmc3k.final.no.umap) + RotatedAxis()
```

```{r DoHeatmap, fig.width=5, fig.height=6}
# DoHeatmap now shows a grouping bar, splitting the heatmap into groups or clusters. This can
# be changed with the `group.by` parameter
DoHeatmap(pbmc3k.final, features = VariableFeatures(pbmc3k.final)[1:100], cells = 1:500, size = 4, angle = 90) + NoLegend()
```

## Applying themes to plots

All plots are ggplot2-based

```{r ggplot customization}
baseplot <- DimPlot(pbmc3k.final, reduction = "umap")
# Add custom labels and titles
baseplot + labs(title = "Clustering of 2,700 PBMCs")

# Seurat also provides several built-in themes, such as DarkTheme; for more details see
# ?SeuratTheme
baseplot + DarkTheme()

# Chain themes together
baseplot + FontSize(x.title = 20, y.title = 20) + NoLegend()
```

## Interactive plotting features

- R plotly graphing library
- works with any ggplot2 scatter (w/ `geom_point`)
- make a ggplot2-based scatter (`DimPlot()` or `FeaturePlot()`) and pass to `HoverLocator()`

```{r Interactive plot}
# Include additional data to display alongside cell names by passing in a data frame of
# information Works well when using FetchData
plot <- FeaturePlot(pbmc3k.final, features = "MS4A1")
HoverLocator(plot = plot, information = FetchData(pbmc3k.final, vars = c("ident", "PC_1", "nFeature_RNA")))
```

Manually select cells for further investigation
- esp for clusters that don't separate unbiasedly, but look distinct
- create ggplo2-based scatter + pass plot to `CellSelector()`
- `CellSelector()` returns vector with names of points selected
- can then set to new identity class + perform differential expression

ex. pretend DCs had merged with monocytes in clustering, but wanted to see what was unique about them based on tSNE position
```{r Manual cell subset}
pbmc3k.final <- RenameIdents(pbmc3k.final, DC = "CD14+ Mono")
plot <- DimPlot(pbmc3k.final, reduction = "umap")
select.cells <- CellSelector(plot = plot)

# Change identity of selection to own mini-cluster
head(select.cells)
Idents(pbmc3k.final, cells = select.cells) <- "NewCells"

# Now, we find markers that are specific to the new cells, and find clear DC markers
newcells.markers <- FindMarkers(pbmc3k.final, ident.1 = "NewCells", ident.2 = "CD14+ Mono", min.diff.pct = 0.3,only.pos = TRUE)
head(newcells.markers)
```
CellSelector to auto assign cell identities
- can return Seurat obj w/ identity classes already set
- pass Seurat object used to make the plot into `CellSelector()` as well as identity class
```{r CellSelector identities}
pbmc3k.final <- CellSelector(plot = plot, object = pbmc3k.final, ident = "selected")
levels(pbmc3k.final)
```

## Plotting Accessories

```{r Label clusters, fig.width=5, fig.height=3}
# LabelClusters and LabelPoints will label clusters (a coloring variable) or individual points
# on a ggplot2-based scatter plot
plot <- DimPlot(pbmc3k.final, reduction = "pca") + NoLegend()
LabelClusters(plot = plot, id = "ident")

# Both functions support `repel`, which will intelligently stagger labels and draw connecting
# lines from the labels to the points or clusters
LabelPoints(plot = plot, points = TopCells(object = pbmc3k.final[["pca"]]), repel = TRUE)
```
Patchwork grid plotting
https://patchwork.data-imaginist.com/

```{r Multple plots in grid, fig.width=5, fig.height=3}
plot1 <- DimPlot(pbmc3k.final)
plot2 <- FeatureScatter(pbmc3k.final, feature1 = "LYZ", feature2 = "CCL5")
# Combine two plots
plot1 + plot2

# Remove the legend from all plots
(plot1 + plot2) & NoLegend()
```

