
# CellMixS: Explore data integration and batch effects

https://www.bioconductor.org/packages/release/bioc/vignettes/CellMixS/inst/doc/CellMixS.html

```{r CellMixS package}
BiocManager::install("CellMixS")
library(CellMixS)
```

`SingleCellExperiment` class

```{r Required packages}
# Load required packages
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(cowplot)
    library(limma)
    library(magrittr)
    library(dplyr)
    library(purrr)
    library(ggplot2)
    library(scater)
})
```

```{r sim50 data}
# Load sim_list example data
sim_list <- readRDS(system.file(file.path("extdata", "sim50.rds"), 
                                package = "CellMixS"))
names(sim_list)
#> [1] "batch0"  "batch20" "batch50"

sce50 <- sim_list[["batch50"]]
class(sce50)
#> [1] "SingleCellExperiment"
#> attr(,"package")
#> [1] "SingleCellExperiment"

table(sce50[["batch"]])
#> 
#>   1   2   3 
#> 250 125 125
```

```{r Visualize batch effect}
# Visualize batch distribution in sce50
visGroup(sce50, group = "batch")
# Visualize batch distribution in other elements of sim_list 
batch_names <- c("batch0", "batch20")
  
vis_batch <- lapply(batch_names, function(name){
    sce <- sim_list[[name]]
    visGroup(sce, "batch") + ggtitle(paste0("sim_", name))
})

plot_grid(plotlist = vis_batch, ncol = 2)
```

```{r Call cms}
# Call cell-specific mixing score for sce50
# Note that cell_min is set to 4 due to the low number of cells and small k.
# Usually default parameters are recommended!! 
sce50 <- cms(sce50, k = 30, group = "batch", res_name = "unaligned", n_dim = 3, cell_min = 4, BPPARAM = BiocParallel::MulticoreParam())
head(colData(sce50))
#> DataFrame with 6 rows and 3 columns
#>           batch cms_smooth.unaligned cms.unaligned
#>        <factor>            <numeric>     <numeric>
#> cell_1        1           0.00000000             0
#> cell_2        1           0.01969896             0
#> cell_3        1           0.00000000             0
#> cell_4        1           0.00823378             0
#> cell_5        1           0.02896438             0
#> cell_6        1           0.06544070             0

# Call cell-specific mixing score for all datasets
sim_list <- lapply(batch_names, function(name){
    sce <- sim_list[[name]]
    sce <- cms(sce, k = 30, group = "batch", res_name = "unaligned", 
               n_dim = 3, cell_min = 4, BPPARAM = BiocParallel::MulticoreParam())
}) %>% set_names(batch_names)

# Append cms50
sim_list[["batch50"]] <- sce50
```

- view as histogram
- cms are p.values from hypothesis testing
- w/o batch effect, p.value histogram should be flat
- increased # of very small p-values == presence of batch-specific bias within data

```{r Visualize cms}
# p-value histogram of cms50
visHist(sce50)
# p-value histogram sim30
# Combine cms results in one matrix
batch_names <- names(sim_list)
cms_mat <- batch_names %>% 
  map(function(name) sim_list[[name]]$cms.unaligned) %>% 
  bind_cols() %>% set_colnames(batch_names)
#> New names:
#> • `` -> `...1`
#> • `` -> `...2`
#> • `` -> `...3`

visHist(cms_mat, n_col = 3)
```

```{r Cell specific visualization}
# cms only cms20
sce20 <- sim_list[["batch20"]]
metric_plot <- visMetric(sce20, metric_var = "cms_smooth.unaligned")

# group only cms20
group_plot <- visGroup(sce20, group = "batch")

plot_grid(metric_plot, group_plot, ncol = 2)
```

```{r Random celltypes}
# Add random celltype assignments as new metadata
sce20[["celltype"]] <- rep(c("CD4+", "CD8+", "CD3"), length.out = ncol(sce20)) %>% 
    as.factor 

visOverview(sce20, "batch", other_var = "celltype")
```

No celltype systemic differences expected w/ random cell types
```{r Random celltype cms}
visCluster(sce20, metric_var = "cms.unaligned", cluster_var = "celltype")
#> Picking joint bandwidth of 0.0996
```
# Evaluate data integration
```{r Run data integration}
# MNN - embeddings are stored in the reducedDims slot of sce
reducedDimNames(sce20)
#> [1] "TSNE" "PCA"  "MNN"
sce20 <- cms(sce20, k = 30, group = "batch", 
             dim_red = "MNN", res_name = "MNN", n_dim = 3, cell_min = 4,
             BPPARAM = BiocParallel::MulticoreParam())

# Run limma
sce20 <- scater::logNormCounts(sce20)
limma_corrected <- removeBatchEffect(logcounts(sce20), batch = sce20$batch)
# Add corrected counts to sce
assay(sce20, "lim_corrected") <- limma_corrected 

# Run cms
sce20 <- cms(sce20, k = 30, group = "batch", 
             assay_name = "lim_corrected", res_name = "limma", n_dim = 3, 
             cell_min = 4,
             BPPARAM = BiocParallel::MulticoreParam())

names(colData(sce20))
#> [1] "batch"                "cms_smooth.unaligned" "cms.unaligned"       
#> [4] "celltype"             "cms_smooth.MNN"       "cms.MNN"             
#> [7] "sizeFactor"           "cms_smooth.limma"     "cms.limma"
```
`limma` + `fastMNN` both flatten p.value distribution, indicating better mixing after batch effect removal
```{r Compare data integration}
# As pvalue histograms
visHist(sce20, metric = "cms.",  n_col = 3)
```

# Remaining batch-specific structure ldfDiff

`ldfDiff`: calc differences btwn each cell's local density factor (ldf) before + after data integration
- ldf: relative measure of cell density around a cell compared to densities within its neighborhood
- calculated on same set of k cells from the cell's knn before integration
- optimal case: relative densities (according to same set of cells) should not change by integration
- `ldfDiff` score should be closs to 0
- overall distribution should be centered around 0 w/o long tails

```{r ldfDiff}
# Prepare input 
# List with single SingleCellExperiment objects 
# (Important: List names need to correspond to batch levels! See ?ldfDiff)
sce_pre_list <- list("1" = sce20[,sce20$batch == "1"], 
                     "2" = sce20[,sce20$batch == "2"], 
                     "3" = sce20[,sce20$batch == "3"])

sce20 <- ldfDiff(sce_pre_list, sce_combined = sce20, 
                 group = "batch", k = 70, dim_red = "PCA", 
                 dim_combined = "MNN", assay_pre = "counts", 
                 n_dim = 3, res_name = "MNN")

sce20 <- ldfDiff(sce_pre_list, sce_combined = sce20, 
                 group = "batch", k = 70, dim_red = "PCA", 
                 dim_combined = "PCA", assay_pre = "counts", 
                 assay_combined = "lim_corrected",  
                 n_dim = 3, res_name = "limma")

names(colData(sce20))
```
- `limma` can preserve batch internal structure within batches
- `fastMNN` not well centered on 0, changes internal structure
  - does not work for batch effect removal on simulated data
  - in line w/ expectations due to small # of genes in example data
  - MNN assumption: batch effects should be much smaller than biological variation
    - not true in small example dataset
```{r visualize ldfDiff}
# ldfDiff score summarized
visIntegration(sce20, metric = "diff_ldf", metric_name = "ldfDiff") 
#> Picking joint bandwidth of 0.0867
```

`evalIntegration`
- `cms`
- `ldfDiff`
- `entropy`
- `mixingMetric`, `localStruct` Seurat equivalents
```{r Test diff metrics}
sce50 <- evalIntegration(metrics = c("isi", "entropy"), sce50, 
                         group = "batch", k = 30, n_dim = 2, cell_min = 4,
                         res_name = c("weighted_isi", "entropy"),
             BPPARAM = BiocParallel::MulticoreParam())
# Visualize with `visMetric`
visOverview(sce50, "batch", 
            metric = c("cms_smooth.unaligned", "weighted_isi", "entropy"), 
            prefix = FALSE)
```





