
# Dynverse

```{r Install dynverse}
# devtools::install_github("dynverse/dyno")
install.packages("hdf5r")
# Docker required - test installation
# Run Docker app first to get the daemon running
dynwrap::test_docker_installation(detailed = TRUE)
```

## Quickstart

https://dynverse.org/users/2-quick_start/

```{r Load libraries}
library(dyno)
library(tidyverse)
```

User guide: preparing the data
https://dynverse.org/users/3-user-guide/1-preparing/

```{r Prepare data}
data("fibroblast_reprogramming_treutlein")

# Sparse matrices
class(fibroblast_reprogramming_treutlein$counts)
class(fibroblast_reprogramming_treutlein$expression)

dataset <- wrap_expression(
  counts = fibroblast_reprogramming_treutlein$counts,
  expression = fibroblast_reprogramming_treutlein$expression
)
```

Uses benchmarking results from dynbenchmark paper.
```{r Select best methods for dataset}
# guidelines <- guidelines_shiny(dataset)
# Reproduces the guidelines as created in the shiny app
answers <- dynguidelines::answer_questions(
  multiple_disconnected = NULL, 
  expect_topology = NULL, 
  expected_topology = NULL, 
  n_cells = 392, 
  n_features = 2000, 
  memory = "2GB", 
  prior_information = c("start_id", "end_id", "end_n", "start_n", "leaves_n", "groups_n", "features_id", "dimred"), 
  docker = TRUE
)
guidelines <- dynguidelines::guidelines(answers = answers)
methods_selected <- guidelines$methods_selected
methods_selected <- c("slingshot", "paga_tree", "grandprix", "paga")
```

```{r Running methods}
model <- infer_trajectory(dataset, first(methods_selected))
```

Several visualization methods to biologically interpret trajectories

ex. combine dimensionality reduction, trajectory model ,and cell clustering
```{r Plotting trajectory}
model <- model %>% add_dimred(dyndimred::dimred_mds, expression_source = dataset$expression)
plot_dimred(
  model, 
  expression_source = dataset$expression, 
  grouping = fibroblast_reprogramming_treutlein$grouping
)
```

```{r Plot gene expression}
plot_dimred(
  model, 
  expression_source = dataset$expression,
  feature_oi = "Fn1"
)
```

```{r Plot groups with background color, fig.width=15, fig.height=10}
plot_dimred(
  model, 
  expression_source = dataset$expression, 
  color_cells = "feature",
  feature_oi = "Vim",
  color_density = "grouping",
  grouping = fibroblast_reprogramming_treutlein$grouping,
  label_milestones = FALSE
)
```

### Interpret trajectory biologically

User guide: adapting the trajectory
https://dynverse.org/users/3-user-guide/5-adapting/

- can use knowledge of start, end, intermediate steps present in data
- use to adapt trajectory so it's easier to interpret

```{r Rooting tree}
model <- model %>% 
  add_root_using_expression(c("Vim"), dataset$expression)
```

- can label milestones w/ marker genes
- then use labels for subsequent analyses + visualization
```{r Milestone labelling}
model <- label_milestones_markers(
  model,
  markers = list(
    MEF = c("Vim"),
    Myocyte = c("Myl1"),
    Neuron = c("Stmn3")
  ),
  dataset$expression
)
```

### Predicting + visualizing genes of interest

```{r Global overview, fig.width=15, fig.height=10}
plot_heatmap(
  model,
  expression_source = dataset$expression,
  grouping = fibroblast_reprogramming_treutlein$grouping,
  features_oi = 50
)
```

```{r Lineage / branch markers, fig.width=15, fig.height=10}
# We can also extract features specific for a branch, eg. genes which change when a cell differentiates into a Neuron

branch_feature_importance <- calculate_branch_feature_importance(
  model,
  expression_source=dataset$expression)

neuron_features <- branch_feature_importance %>% 
  filter(to == names(model$milestone_labelling)[which(model$milestone_labelling =="Neuron")]) %>% 
  top_n(50, importance) %>% 
  pull(feature_id)

plot_heatmap(
  model, 
  expression_source = dataset$expression, 
  features_oi = neuron_features
)
```

```{r Genes important at bifurcation points, fig.width=15, fig.height=10}
branching_milestone <- model$milestone_network %>%
  group_by(from) %>%
  filter(n() > 1) %>%
  pull(from) %>%
  first()

branch_feature_importance <- calculate_branching_point_feature_importance(
  model,
  expression_source = dataset$expression,
  milestones_oi = branching_milestone)

branching_point_features <- branch_feature_importance %>%
  top_n(20, importance) %>%
  pull(feature_id)

plot_heatmap(
  model,
  expression_source = dataset$expression,
  features_oi = branching_point_features
)
```

```{r Milestone genes visualization, fig.width=15, fig.height=12}
space <- dyndimred::dimred_mds(dataset$expression)

map(branching_point_features[1:12], function(feature_oi) {
  plot_dimred(
    model,
    dimred = space,
    expression_source = dataset$expression,
    feature_oi = feature_oi,
    label_milestones = FALSE) +
    theme(legend.position = "none") +
    ggtitle(feature_oi)
}) %>% patchwork::wrap_plots()
```

