
# Seurat: Differential Expression Testing

https://satijalab.org/seurat/articles/de_vignette.html

```{r Load data}
library(Seurat)
library(SeuratData)
pbmc <- LoadData("pbmc3k", type = "pbmc3k.final")
```

# Default DE tests

`FindMarkers()` - Seurat DE features
- non-parametric Wilcoxon rank sum test
- for 2 groups of cells, specify `ident.1` & `ident.2`
- if `ident.2` non specified or NULL, DE test btwn group and all others

```{r FindMarkers}
# list options for groups to perform differential expression on
levels(pbmc)
# Find differentially expressed features between CD14+ and FCGR3A+ Monocytes
monocyte.de.markers <- FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono")
# view results
head(monocyte.de.markers)
```

Results data frame columns:
- `p_val`: unadjusted p value
- `avg_log2FC`: log fold change of avg expression btwn two groups
  - positive values = feature more highly expressed in first group
- `pct.1`: % of cells where feature detected in 1st group
- `pct.2`: % of cells where feature detected in 2nd group
- `p_val_adj`: adjusted pval, based on Bonferroni correction using all features in dataset

```{r DE one vs all others}
# Find differentially expressed features between CD14+ Monocytes and all other cells, only
# search for positive markers
monocyte.de.markers <- FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = NULL, only.pos = TRUE)
# view results
head(monocyte.de.markers)
```

# Prefilter features / cells

- increase speed of marker discovery
- features infrequently detected in either group of cells or features expressed at similar avg levels unlikely to be DE
- ex. `min.pct`, `logfc.threshold`, `min.diff.pct`, `max.cells.per.ident`

```{r Prefilter min.pct}
# Pre-filter features that are detected at <50% frequency in either CD14+ Monocytes or FCGR3A+
# Monocytes
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", min.pct = 0.5))
```

```{r Prefilter logfc.threshold}
# Pre-filter features that have less than a two-fold change between the average expression of
# CD14+ Monocytes vs FCGR3A+ Monocytes
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", logfc.threshold = log(2)))
```

```{r Prefilter min.diff.pct}
# Pre-filter features whose detection percentages across the two groups are similar (within
# 0.25)
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", min.diff.pct = 0.25))
```

```{r Prefilter max.cells.per.ident (subsampling)}
# Increasing min.pct, logfc.threshold, and min.diff.pct, will increase the speed of DE
# testing, but could also miss features that are prefiltered

# Subsample each group to a maximum of 200 cells. Can be very useful for large clusters, or
# computationally-intensive DE tests
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", max.cells.per.ident = 200))
```

# DE w/ alternative tests

Supported DE tests
- wilcox : Wilcoxon rank sum test (default)
- bimod : Likelihood-ratio test for single cell feature expression, (McDavid et al., Bioinformatics, 2013)
- roc : Standard AUC classifier
- t : Student’s t-test
- poisson : Likelihood ratio test assuming an underlying negative binomial distribution. Use only for UMI-based datasets
- negbinom : Likelihood ratio test assuming an underlying negative binomial distribution. Use only for UMI-based datasets
- LR : Uses a logistic regression framework to determine differentially expressed genes. Constructs a logistic regression model predicting group membership based on each feature individually and compares this to a null model with a likelihood ratio test.
- MAST : GLM-framework that treates cellular detection rate as a covariate (Finak et al, Genome Biology, 2015)
- DESeq2 : DE based on a model using the negative binomial distribution (Love et al, Genome Biology, 2014)

`test.use` parameter specifies which DE test

```{r MAST DE}
# Test for DE features using the MAST package
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", test.use = "MAST"))
```

```{r DESeq2 DE}
# Test for DE features using the DESeq2 package. Throws an error if DESeq2 has not already been installed
# Note that the DESeq2 workflows can be computationally intensive for large datasets, but are incompatible with some feature pre-filtering options We therefore suggest initially limiting the number of cells used for testing
head(FindMarkers(pbmc, ident.1 = "CD14+ Mono", ident.2 = "FCGR3A+ Mono", test.use = "DESeq2", max.cells.per.ident = 50))
```



