
# "Using fgsea package" Vignette

https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html

```{r Load libraries}
library(fgsea)
library(data.table)
library(ggplot2)
```

## Quick run
```{r Load example pathways/ranks + random seed}
data(examplePathways)
data(exampleRanks)
set.seed(42)
```

```{r Run fgsea}
fgseaRes <- fgsea(pathways = examplePathways, 
                  stats    = exampleRanks,
                  minSize  = 15,
                  maxSize  = 500)
```

Resulting table has enrichment scores and p-values
```{r See results}
head(fgseaRes[order(pval), ])
```

`fgsea` has default lower bound `eps=1e-00` for estimating P values
- Estimate P-value more accurately -> set `eps` argument to zero in function
```{r Finer p-value estimation}
fgseaRes <- fgsea(pathways = examplePathways, 
                  stats    = exampleRanks,
                  eps      = 0.0,
                  minSize  = 15,
                  maxSize  = 500)

head(fgseaRes[order(pval), ])
```

```{r Enrichment plot for pathway}
plotEnrichment(examplePathways[["5991130_Programmed_Cell_Death"]],
               exampleRanks) + labs(title="Programmed Cell Death")
```

Using as_ggplot workaround to actually output the plot..
- https://github.com/ctlab/fgsea/issues/101
```{r Table plot for selected pathways}
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
as_ggplot(plotGseaTable(examplePathways[topPathways], exampleRanks, fgseaRes, 
              gseaParam=0.5, render = F))
```
```{r Collapse similar pathways}
collapsedPathways <- collapsePathways(fgseaRes[order(pval)][padj < 0.01], 
                                      examplePathways, exampleRanks)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway]
as_ggplot(plotGseaTable(examplePathways[mainPathways], exampleRanks, fgseaRes, 
              gseaParam = 0.5, render = F))
```

```{r Save as text format}
# data.table::fwrite(fgseaRes, file="fgseaRes.txt", sep="\t", sep2=c("", " ", ""))
```

Similar to `AnnotationDbi::mapIds` + corresponding database
```{r Make leading edge more human-readable}
library(org.Mm.eg.db)
fgseaResMain <- fgseaRes[match(mainPathways, pathway)]
fgseaResMain[, leadingEdge := mapIdsList(
                                     x=org.Mm.eg.db, # Mouse db for this example
                                     keys=leadingEdge,
                                     keytype="ENTREZID", 
                                     column="SYMBOL")]
# fwrite(fgseaResMain, file="fgseaResMain.txt", sep="\t", sep2=c("", " ", ""))
```

## Using Reactome pathways
```{r Reactome fgsea}
library(reactome.db)
pathways <- reactomePathways(names(exampleRanks))
fgseaRes <- fgsea(pathways, exampleRanks, maxSize=500)
head(fgseaRes)
```

## Starting from files (like GSEA)
```{r Load example rank + gmt}
rnk.file <- system.file("extdata", "naive.vs.th1.rnk", package="fgsea")
gmt.file <- system.file("extdata", "mouse.reactome.gmt", package="fgsea")
# Load ranks
ranks <- read.table(rnk.file,
                    header=TRUE, colClasses = c("character", "numeric"))
ranks <- setNames(ranks$t, ranks$ID)
str(ranks)
# Load pathways (gmt file?)
pathways <- gmtPathways(gmt.file)
str(head(pathways))
```

```{r Run fgsea}
fgseaRes <- fgsea(pathways, ranks, minSize=15, maxSize=500)
head(fgseaRes)
```
